Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-04-19 paper.tex

@@

\rljf{Small formatting changes I made: \\
* I prefer bold over italics when introducing terms, because when someone is visually scanning the paper looking for a definition, bold stands out much better. \\
* Punctuation should be attached to equations as if they were part of the text (so sometimes periods and commas after equations) \\
* no spaces between m-dashes---like this \\
Also, something crazy is going on with the line numbers!
}

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Estimation (OCE, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts $(X_{-M \dots -1}), (X_{0 \dots M-1})$:
\begin{equation}\label{eq:ib-oce}
	\max_{Z} \left( \operatorname{I}[X_{-M\dots -1}, Z] - \lambda \cdot \operatorname{I}[Z, \finitefuture] \right).
\end{equation}
Second, it estimates information measures directly from the observed counts of $(X_{-M\dots -1}), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|X_{-M \dots -1})$ that maps each observed past sequence $X_{-M...-1}$ to a distribution over a (finite) set of code words $Z$.
%From this, one obtains a corresponding decoder $P(X_{0...M}|Z)$ via Bayes' Rule.

@@

As in OCE (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: \rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read.}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | X_{-M\dots-1}} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[X_{-M\dots -1}, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | X_{-M\dots -1}
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | X_{-M\dots-1}}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ X_{-M\dots-1}, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

When deploying such a predictive code $Z$, two components have to be computed: a distribution $P(Z|X_{-M\dots -1})$ that \key{encodes} past observations into a code $Z$, and a distribution $P(\finitefuture|Z)$ that \key{decodes} the code $Z$ into predictions about the future.

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|X_{-M\dots -1}),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $X_{-M\dots -1}$ into the parameters of the distribution $P_\phi(\cdot|X_{-M\dots -1})$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P(\finitefuture, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(\finitefuture|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[X_{-M\dots -1}, Z]$ of the code $Z \sim P_\phi(Z|x_{-M\dots -1})$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

\begin{proposition}
The loss incurred when predicting the future from $Z$ via $\psi$ upper-bounds the true conditional entropy of the future given $Z$, when predicting using the exact decoding distribution (\ref{eq:cond-fut}):
\begin{equation}\label{ineq1}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(\finitefuture | Z)\right] \geq \operatorname{H}[\finitefuture|Z].
\end{equation}
Furthermore, equality is attained if and only if $P_\psi(\finitefuture|Z) = P(\finitefuture|Z)$. % (as defined in~(\ref{eq:cond-fut})).
\end{proposition}

@@

\begin{proposition} 
The KL Divergence between $P_\phi(Z|X_{-M\dots-1})$ and $q(Z)$, averaged over pasts $X_{-M\dots-1}$, upper-bounds the rate of $Z$: % mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X_{-M\dots-1}}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots-1})}{q(Z)}\right] &=   \mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log \frac{P_\phi(Z|X_{-M\dots -1})}{q(Z)}  \\
	& \geq  \mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log \frac{P_\phi(Z|X_{-M\dots -1})}{P(Z)}  \\
	& = \operatorname{I}[X_{-M\dots -1},Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})$.
\end{proposition}

@@

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction loss for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[\finitefuture|X_{-M\dots-1}]
  \operatorname{H}[\finitefuture|Z] \leq   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(\finitefuture | Z)\right]. % \operatorname{H}[\finitefuture]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and---if the family of approximators is sufficiently rich---can come close to turning these into the equalities (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

@@

Following standard practice in machine learning, we split the data set of sample time series into three partitions (\emph{training set}, \emph{validation set}, and \emph{test set}). \rljf{what proportions}
We use the training set for optimizing the parameters as described above.
After every pass through the training set, the objective (\ref{eq:bound}) is evaluated on the validation set using a Monte Carlo estimate with one sample $Z$ per trajectory; optimization terminates once the value on the held-out set does not decrease any more.
%After this, we use the test set to estimate rate and predictiveness of the fully optimized model (see next paragraph).

%\paragraph{Evaluation}
After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction loss on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction loss (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1}$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots -1})}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)},
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $X_{-M\dots -1}$ as in (\ref{eq:z-gauss}), and prediction loss 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(\finitefuture | Z)\right] \approx - \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(\finitefuture | Z).
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction loss achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

\paragraph{OCE}
As discussed in Section~\ref{sec:oce}, OCE is affected by overfitting, %overestimates structure in the process due to overfitting, %resulting in biased estimates of rate and distortion.More specifically, 
%overfitting will result in an overly optimistic estimate, 
and will systematically overestimate the predictiveness achieved at a given rate.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and predictiveness on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|X_{-M\dots -1})$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCE and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{ineq2}:
\begin{equation}\label{eq:rate-oce}
\frac{1}{N} \sum_{X_{-M}\dots X_{-1} \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|X_{-M}\dots X_{-1})}{s(Z)}
\end{equation}
where $P(Z|X_{-M}\dots X_{-1})$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-oce}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{ineq1}:
\begin{equation}
\frac{1}{N} \sum_{X_{-M}\dots X_{M-1} \in \text{Test Data}} \E_{Z \sim P(\cdot|X_{-M \dots -1})} \log P(\finitefuture|Z).
\end{equation}
where $P(\finitefuture|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to (\ref{ineq1}-\ref{ineq2}), these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\paragraph{NPRD}
For all experiments, we used $M=15$.
Neural networks have hyperparameters, such as the number of units and the step size used in optimization, which affect the quality of approximation depending on properties of the dataset and the function being approximated.
Given that NPRD provably provides upper bounds on the PRD objective~(\ref{eq:prd-recall}), one can in principle identify the best hyperparameters for a given process by choosing the combination that leads to the lowest estimated upper bounds.
As a computationally more efficient method, we defined a range of plausible hyperparameters based both on experience reported in the literature, and considerations of computational efficiency.
These parameters are shown in Table~\ref{tab:nprd-hyperparameters}.
We then randomly sampled, for each of the processes that we experimented on, combinations of $\lambda$ and these hyperparameters to run NPRD on.

@@

We first test NPRD on two processes where the predictive rate-distortion tradeoff is analytically tractable. % by computing the epsilon machine \citep{marzen-predictive-2016}. %, and where prior work has found the epsilon machine.
The \emph{Even Process} \citep{marzen-predictive-2016} is the process of 0/1 IID coin flips, conditioned on all blocks of consecutive ones having even length.
Its complexity and excess entropy are both $\approx 0.63$ nats. % (stationary distribution: 1/3 for odd, 2/3 for even)
It has infinite Markov order, and \cite{marzen-predictive-2016} find that OCE (at $M=5$) performs poorly. %, overestimating  even at $M=5$.
The true predictive rate-distortion curve was computed in \cite{marzen-predictive-2016} using the analytically known epsilon machine.
The \emph{Random Insertion Process} \citep{marzen-predictive-2016} consists of sequences of uniform coin flips $X_t \in \{0,1\}$, subject to the constraint that, if $X_{t-2}$ was a 0, then $X_{t}$ has to be a 1.

We applied NPRD to these processes by training on 3M random trajectories of length 30, and using 3000 additional trajectories for validation and test data.
For each process, we ran NPRD 1000 times for randomly choices of $\lambda \in [0,1]$.
Due to computational constraints, when running OCE, we limited sample size to 3000 trajectories for estimation and as held-out data.
Following \cite{marzen-predictive-2016}, we ran OCE for $M=1,...,5$.

The resulting estimates are shown in Figure~\ref{fig:even}, together with the analytical rate-distortion curves computed by \cite{marzen-predictive-2016}.
Individual runs of NPRD show variation (red dots), but most runs lead to results close to the analytical curve (gray line), and strongly surpass the curves computed by OCE at $M=5$.
Bounding the tradeoff curve using the sets of runs of NPRD results in a close fit (red line) to the analytical tradeoff curves.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{code/figures/even-info.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/rip-info.pdf}

@@

\paragraph{Recovering Causal States}
Does NPRD lead to interpretable codes $Z$?
To answer this, we further investigated the NPRD approximation to the Random Insertion Process (RIP), obtained in the previous paragraph.
The $\epsilon$-machine was computed by  \cite{marzen-predictive-2016} and is given in Figure~\ref{fig:rip-machine} (left).
The process has three causal states:
State \textit{A} represents those pasts where the future starts with $1^{2k}0$ ($k = 0, 1, 2, \dots$) -- these are the pasts ending in either $001$ or $10111^m$ ($m= 0, 1, 2, \dots$). % but cannot go in with 10
State \textit{B} represents those pasts ending in 10 -- the future has to start with 01 or 11.
State \textit{C} represents those pasts ending in either 00 or 01 -- the future has to start with $1^{2k+1}0$ ($k = 0, 1, 2, \dots$).

The analytical solution to the predictive rate-distortion problem was computed by \cite{marzen-predictive-2016}.
At $\lambda > 0.5$, the optimal solution collapses $A$ and $B$ into a single codeword, while all three states are mapped to separate codewords for $\lambda \leq 0.5$.

>>>>

\let\oldequation\equation
\let\oldendequation\endequation

\renewenvironment{equation}
  {\linenomathNonumbers\oldequation}
  {\oldendequation\endlinenomath}

@@

\rljf{Small formatting changes I made: \\
* I prefer bold over italics when introducing terms, because when someone is visually scanning the paper looking for a definition, bold stands out much better. \\
* Punctuation should be attached to equations as if they were part of the text (so sometimes periods and commas after equations) \\
* no spaces between m-dashes---like this \\
* we should make sure not to use ``this'', ``that'', ``which'', etc. as bare NPs. Copyeditors don't like this.
}

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Estimation (OCE, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts $(X_{-M \dots -1}), (\finitefuture)$:
\begin{equation}\label{eq:ib-oce}
	\max_{Z} \left( \operatorname{I}[\finitepast, Z] - \lambda \cdot \operatorname{I}[Z, \finitefuture] \right).
\end{equation}
Second, it estimates information measures directly from the observed counts of $(\finitepast), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|X_{-M \dots -1})$ that maps each observed past sequence $X_{-M...-1}$ to a distribution over a (finite) set of code words $Z$.
%From this, one obtains a corresponding decoder $P(X_{0...M}|Z)$ via Bayes' Rule.

@@

As in OCE (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: \rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read. $^M\overleftarrow{X}$}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | X_{-M\dots-1}} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | X_{-M\dots-1}}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ X_{-M\dots-1}, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

When deploying such a predictive code $Z$, two components have to be computed: a distribution $P(Z|\finitepast)$ that \key{encodes} past observations into a code $Z$, and a distribution $P(\finitefuture|Z)$ that \key{decodes} the code $Z$ into predictions about the future.

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|\finitepast),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $\finitepast$ into the parameters of the distribution $P_\phi(\cdot|\finitepast)$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P(\finitefuture, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(\finitefuture|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[\finitepast, Z]$ of the code $Z \sim P_\phi(Z|\finitepast)$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

\begin{proposition}
\label{prop:mi-bound}
The loss incurred when predicting the future from $Z$ via $\psi$ upper-bounds the true conditional entropy of the future given $Z$, when predicting using the exact decoding distribution (\ref{eq:cond-fut}):
\begin{equation}\label{ineq1}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(\finitefuture | Z)\right] \geq \operatorname{H}[\finitefuture|Z].
\end{equation}
Furthermore, equality is attained if and only if $P_\psi(\finitefuture|Z) = P(\finitefuture|Z)$. % (as defined in~(\ref{eq:cond-fut})).
\end{proposition}

@@

\begin{proposition} 
\label{prop:kl-bound}
The KL Divergence between $P_\phi(Z|X_{-M\dots-1})$ and $q(Z)$, averaged over pasts $X_{-M\dots-1}$, upper-bounds the rate of $Z$: % mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X_{-M\dots-1}}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|X_{-M\dots-1})}{q(Z)}\right] &=   \mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log \frac{P_\phi(Z|\finitepast)}{q(Z)}  \\
	& \geq  \mathbb{E}_{X_{-M\dots-1}} \mathbb{E}_{Z|X_{-M\dots-1}} \log \frac{P_\phi(Z|\finitepast)}{P(Z)}  \\
	& = \operatorname{I}[\finitepast,Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})$.
\end{proposition}

@@

If $\phi, \psi, q$ are not exact solutions to (\ref{eq:bound}), the two propositions guarantee that we still have bounds on rate and prediction loss for the code $Z$ generated by $\phi$:
\begin{equation}\label{eq:bound1-new}
 \operatorname{I}[Z, X_{-M\dots-1}] \leq   \mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
\begin{equation}\label{eq:bound2-new}
  %I[X_{-M\dots 0}, X_{0\dots M} | Z] \leq -	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(X_{0\dots M} | z)\right]- H[\finitefuture|X_{-M\dots-1}]
  \operatorname{H}[\finitefuture|Z] \leq   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(X_{-M\dots-1})}\left[\log P_\psi(\finitefuture | Z)\right]. % \operatorname{H}[\finitefuture]
\end{equation}
To carry out the optimization~(\ref{eq:bound}), we will restrict $\phi, \psi, q$ to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
While the solutions may not be exact solutions to the nonparametric objective~(\ref{eq:bound}), they will still satisfy the bounds (\ref{eq:bound1-new}-\ref{eq:bound2-new}), and---if the family of approximators is sufficiently rich---can come close to turning these into the equalities (\ref{eq:eq1-new}-\ref{eq:eq2-new}).

@@

Following standard practice in machine learning, we split the data set of sample time series into three partitions (\emph{training set}, \emph{validation set}, and \emph{test set}). 
We use the training set for optimizing the parameters as described above.
After every pass through the training set, the objective (\ref{eq:bound}) is evaluated on the validation set using a Monte Carlo estimate with one sample $Z$ per trajectory; optimization terminates once the value on the held-out set does not decrease any more.
%After this, we use the test set to estimate rate and predictiveness of the fully optimized model (see next paragraph).

%\paragraph{Evaluation}
After optimizing the parameters on a set of observed trajectories, we estimate rate and prediction loss on the test set.
Given parameters for $\phi, \psi, q$, we evaluate the PRD objective~(\ref{eq:bound}), rate (\ref{eq:bound1-new}), and the prediction loss (\ref{eq:bound2-new}) on the test set by taking, for each time series $X_{-M}...X_{M-1}$ from the test set, a single sample $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing Monte Carlo estimates for rate
\begin{equation}\label{eq:bound-mc-rate}
\mathbb{E}_{X}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right] \approx	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}  \log \frac{p_{\mathcal{N}(\mu, \sigma^2)}(z)}{q(z)},
\end{equation}
where $p_{\mathcal{N}(\mu, \sigma^2)}(z)$ is the Gaussian density with $\mu, \sigma^2$ computed from $\finitepast$ as in (\ref{eq:z-gauss}), and prediction loss 
\begin{equation}\label{eq:bound-mc-distortion}
 -	\mathbb{E}_{Z \sim \phi(X_{-M\dots-1})}\left[\log P_\psi(\finitefuture | Z)\right] \approx - \frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P_\psi(\finitefuture | Z).
\end{equation}
%and the PRD objective
%\begin{equation}\label{eq:bound-mc}
%	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
%\end{equation}
%Note that one can go from distortion to predictiveness by estimating
Thanks to (\ref{eq:bound1-new}-\ref{eq:bound2-new}), these estimates are guaranteed to be \emph{upper bounds} on the true rate and prediction loss achieved by the code $Z  \sim \mathcal{N}(\mu, \sigma^2)$, up to sampling error introduced into the Monte Carlo estimation by sampling $z$ and the finite size of the test set.

@@

\paragraph{OCE} As discussed in Section~\ref{sec:oce}, OCE is affected by overfitting, %overestimates structure in the process due to overfitting, %resulting in biased estimates of rate and distortion.More specifically, 
%overfitting will result in an overly optimistic estimate, 
and will systematically overestimate the predictiveness achieved at a given rate.
To address this problem, we follow the evaluation method used for NPRD, evaluating rate and predictiveness on held-out test data.
We partition the available time series data into a training and test set. We use the training set to create the encoder $P(Z|\finitepast)$ using the Blahut-Arimoto algorithm as described by~\citet{still-optimal-2010}.
We then use the held-out test set to estimate rate and prediction loss.
This method not only enables fair comparison between OCE and NPRD, but also provides a more realistic evaluation, by focusing on the performance of the code $Z$ when deployed on new data samples.
For rate, we use the same variational bound that we use for NPRD, stated in Proposition~\ref{prop:kl-bound}:
\begin{equation}\label{eq:rate-oce}
\frac{1}{N} \sum_{X_{-M}\dots X_{-1} \in \text{Test Data}} \operatorname{D_{KL}}\infdivx{P(Z|X_{-M}\dots X_{-1})}{s(Z)}
\end{equation}
where $P(Z|X_{-M}\dots X_{-1})$ is the encoder created by the Blahut-Arimoto algorithm, and $s(Z)$ is the marginal distribution of $Z$ on the training set.
$N$ is the number of sample time series in the test data.
In the limit of enough training data, when $s(Z)$ matches the actual population marginal of $Z$, (\ref{eq:rate-oce}) is an unbiased estimate of the rate. %  If the training set is large enough and 
%\textbf{Is $s$ defined anywhere? same for $r$?}
We estimate the prediction loss on the future observations as the empirical cross-entropy, i.e., the variational bound stated in Proposition~\ref{prop:mi-bound}:
\begin{equation}
\frac{1}{N} \sum_{X_{-M}\dots X_{M-1} \in \text{Test Data}} \E_{Z \sim P(\cdot|X_{-M \dots -1})} \log P(\finitefuture|Z).
\end{equation}
where $P(\finitefuture|Z)$ is the decoder obtained from the Blahut-Arimoto algorithm on the training set.
Thanks to Propositions~\ref{prop:mi-bound} and \ref{prop:kl-bound}, these quantities provide upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, sampling error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

@@

\paragraph{NPRD}
For all experiments, we used $M=15$.
Neural networks have hyperparameters, such as the number of units and the step size used in optimization, which affect the quality of approximation depending on properties of the dataset and the function being approximated.
Given that NPRD provably provides upper bounds on the PRD objective~(\ref{eq:prd-loss}), one can in principle identify the best hyperparameters for a given process by choosing the combination that leads to the lowest estimated upper bounds.
As a computationally more efficient method, we defined a range of plausible hyperparameters based both on experience reported in the literature, and considerations of computational efficiency.
These parameters are shown in Table~\ref{tab:nprd-hyperparameters}.
We then randomly sampled, for each of the processes that we experimented on, combinations of $\lambda$ and these hyperparameters to run NPRD on.

@@

We first test NPRD on two processes where the predictive rate-distortion tradeoff is analytically tractable. % by computing the epsilon machine \citep{marzen-predictive-2016}. %, and where prior work has found the epsilon machine.
The \key{Even Process} \citep{marzen-predictive-2016} is the process of 0/1 IID coin flips, conditioned on all blocks of consecutive ones having even length.
Its complexity and excess entropy are both $\approx 0.63$ nats. % (stationary distribution: 1/3 for odd, 2/3 for even)
It has infinite Markov order, and \citet{marzen-predictive-2016} find that OCE (at $M=5$) performs poorly. %, overestimating  even at $M=5$.
The true predictive rate-distortion curve was computed in \cite{marzen-predictive-2016} using the analytically known epsilon machine.
The \key{Random Insertion Process} \citep{marzen-predictive-2016} consists of sequences of uniform coin flips $X_t \in \{0,1\}$, subject to the constraint that, if $X_{t-2}$ was a 0, then $X_{t}$ has to be a 1.

We applied NPRD to these processes by training on 3M random trajectories of length 30, and using 3000 additional trajectories for validation and test data.
For each process, we ran NPRD 1000 times for randomly choices of $\lambda \in [0,1]$.
Due to computational constraints, when running OCE, we limited sample size to 3000 trajectories for estimation and as held-out data.
Following \citet{marzen-predictive-2016}, we ran OCE for $M=1,...,5$.

The resulting estimates are shown in Figure~\ref{fig:even}, together with the analytical rate-distortion curves computed by \citet{marzen-predictive-2016}.
Individual runs of NPRD show variation (red dots), but most runs lead to results close to the analytical curve (gray line), and strongly surpass the curves computed by OCE at $M=5$.
Bounding the tradeoff curve using the sets of runs of NPRD results in a close fit (red line) to the analytical tradeoff curves.

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{code/figures/even-info.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/rip-info.pdf}

@@

\paragraph{Recovering Causal States}
Does NPRD lead to interpretable codes $Z$?
To answer this, we further investigated the NPRD approximation to the Random Insertion Process (RIP), obtained in the previous paragraph.
The $\epsilon$-machine was computed by  \citet{marzen-predictive-2016} and is given in Figure~\ref{fig:rip-machine} (left).
The process has three causal states:
State \textit{A} represents those pasts where the future starts with $1^{2k}0$ ($k = 0, 1, 2, \dots$) -- these are the pasts ending in either $001$ or $10111^m$ ($m= 0, 1, 2, \dots$). % but cannot go in with 10
State \textit{B} represents those pasts ending in 10---the future has to start with 01 or 11.
State \textit{C} represents those pasts ending in either 00 or 01---the future has to start with $1^{2k+1}0$ ($k = 0, 1, 2, \dots$).

The analytical solution to the predictive rate-distortion problem was computed by \citet{marzen-predictive-2016}.
At $\lambda > 0.5$, the optimal solution collapses $A$ and $B$ into a single codeword, while all three states are mapped to separate codewords for $\lambda \leq 0.5$.
